### 2023-10-27 calculus.tex
Clean up references.bib

The key contribution of this section is the \emph{marked lambda calculus}, a calculus based on the
gradually typed lambda calculus (GTLC) that formalizes the mechanism by which errors like these can
be localized, and how to recover, in all cases, from such errors. 
% In brief, the marked lambda calculus formalizes bidirectional type error localization with gradual
% recovery. 
\Cref{sec:calculus-calculus} introduces the syntax, judgemental structure, guiding metatheory, and
then goes through the rules, organized by syntactic form rather than judgement, to build intuition
about how the various judgements relate to one another.
All rules and theorems may be found organized by judgement form in the supplementary appendix for
reference, alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis}
(discussed in \cref{sec:calculus-agda}).
We intentionally keep the marked lambda calculus minimal, because it is intended to capture the
essential idea and introduce a general pattern that language designers can employ to created marked
variants of their own type systems.
As an initial example, we consider a surprisingly subtle combination of features as a more
substantial case study: the combination of destructing let expressions with granular type
annotations, in \cref{sec:calculus-let}.
Finally, \cref{sec:calculus-polymorphism} briefly explores how the system might be extended to more
complex judgemental structures, such as parametric polymorphism.

@@

The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
with numbers, booleans, pairs, and empty expression holes \cite{HazelnutSNAPL}.
Given in \cref{fig:calculus-syntax}, the syntax consists of two expression languages:
%
\begin{itemize}
  \item The \emph{unmarked language}, which is the original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

@@

$\TUnknown$ is a type hole, which we identify with the unknown type from gradual type theory \cite{Siek06a}.
Finally, to model the edit state of a program in development, $\EEHole$ denotes an \emph{empty expression hole},
used to represent syntactically incomplete portions of the program à la Hazelnut
\cite{HazelnutPOPL}. Note, however, that these empty expression holes are not at all semantically
critical to the calculus---they are included illustrate how the calculus fits into the larger
problem of modelling incomplete program states and to discuss polymorphic generalization in
\cref{sec:polymorphicGlobal}.

@@

As our starting point, types classify unmarked expressions by a completely standard \emph{bidirectional type system} \cite{BidirTyping,Localinf}, which employs
two mutually defined judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
establishes that, under the typing context $\ctx$, the expression $e$ synthesizes or locally infers the type $\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$,
states that the expression $\EMV$ may appear where an expression of type $\TMV$ is expected.

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. Analysis employs the judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$,
which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini:2016}, i.e. it may be
considered an arrow type. Defined in \cref{fig:calculus-matched-arrow}, this notion is purely a
technical mechanism to avoid duplication of rules related to arrow types \cite{siek2015refined}.

@@

We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of Hazel \cite{hazel},
a live functional programming environment in which incomplete programs, i.e., programs that contain holes, may be type-checked, manipulated, and even executed.
In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above).
Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser.
The supplemental material contains a pre-built implementation.

The result of this effort is the first full-scale typed functional language equipped with a language server that solves the semantic gap problem: all of Hazel's semantic services are available as long as the program sketch is syntactically well-formed, which include type hints \cite{potter2020hazel}, semantic navigation, semantic code completion \cite{potter2020hazel,blinn2022integrative}, contextualized documentation \cite{potter2022contextualized}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelLive}, even semantic services that require run-time information, like testing.
%

Hazel offers both a textual syntax, with explicit empty holes, and more interestingly,
a structure editor that is able to additionally offer partial syntactic error recovery.
In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automatically insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{moon2022tylr,DBLP:conf/vl/Moon023}. Efforts outside of the scope of this paper are underway to define syntactic error recovery mechanisms that can handle unmatched delimiters as well, which would achieve total syntax error recovery. In combination with our contributions, this would achieve the Hazel project's stated goal of ensuring that \emph{there are no meaningless editor states}.

@@

Earlier versions of Hazel achieved total type error recovery by using a term-based structure editor, which maintained delimiter matching by construction, albeit at the cost of some syntactic rigidity.
To specify such a structure editor formally, \citet{HazelnutPOPL} introduced the Hazelnut action calculus, which operates over a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor}.
The synthetic action judgment, written $\ASEAction{\ctx}{\ZMV}{\TMV}{\ZMV'}{\TMV'}{\AMV}$, performs some edit action $\AMV$ on some \emph{zippered expression} $\ZMV$ (which is an expression with a superimposed cursor, denoted $\hat{\EMV}$ in the original work) that synthesizes type $\TMV$, producing some new $\ZMV'$ that synthesizes types $\TMV'$.
Analogously, the analytic action judgment is written $\AAEAction{\ctx}{\ZMV}{\ZMV'}{\TMV}{\AMV}$.
Type and binding information is propagated to the location of the cursor during these operations, allowing them to insert the necessary error marks (i.e. non-empty holes, in Hazelnut's parlance).
When, for example, wrapping an expression whose type is inconsistent with $\TNum$ in an arithmetic operation, the calculus inserts an error mark around that operand:
%
\begin{mathpar}
    \inferrule{
        \inconsistent{\TMV}{\TNum}
    }{
        \ASCon{\ctx}{\ZCursor{\EMV}}{\TMV}{\ZPlusL{\ECInconType{\ECMV}}{\ZCursor{\EEHole}}}{\TNum}{\SPlusL}
    }
\end{mathpar}

>>>>

The key contribution of this section is the \emph{marked lambda calculus}, a calculus based on the
gradually typed lambda calculus (GTLC) that formalizes the mechanism by which errors like these can
be localized, and how to recover, in all cases, from such errors. 
% In brief, the marked lambda calculus formalizes bidirectional type error localization with gradual
% recovery. 
\Cref{sec:calculus-calculus} introduces the syntax, judgemental structure, guiding metatheory, and
then goes through the rules, organized by syntactic form rather than judgement, to build intuition
about how the various judgements relate to one another.
All rules and theorems may be found organized by judgement form in the supplementary appendix for
reference, alongside a complete mechanization in the Agda proof assistant \cite{norell2007}
(discussed in \cref{sec:calculus-agda}).
We intentionally keep the marked lambda calculus minimal, because it is intended to capture the
essential idea and introduce a general pattern that language designers can employ to created marked
variants of their own type systems.
As an initial example, we consider a surprisingly subtle combination of features as a more
substantial case study: the combination of destructing let expressions with granular type
annotations, in \cref{sec:calculus-let}.
Finally, \cref{sec:calculus-polymorphism} briefly explores how the system might be extended to more
complex judgemental structures, such as parametric polymorphism.

@@

The marked lambda calculus is based on the gradually typed lambda calculus \cite{siek2006} extended
with numbers, booleans, pairs, and empty expression holes \cite{omar2017a}.
Given in \cref{fig:calculus-syntax}, the syntax consists of two expression languages:
%
\begin{itemize}
  \item The \emph{unmarked language}, which is the original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

@@

$\TUnknown$ is a type hole, which we identify with the unknown type from gradual type theory \cite{siek2006}.
Finally, to model the edit state of a program in development, $\EEHole$ denotes an \emph{empty expression hole},
used to represent syntactically incomplete portions of the program à la Hazelnut
\cite{omar2017b}. Note, however, that these empty expression holes are not at all semantically
critical to the calculus---they are included illustrate how the calculus fits into the larger
problem of modelling incomplete program states and to discuss polymorphic generalization in
\cref{sec:polymorphicGlobal}.

@@

As our starting point, types classify unmarked expressions by a completely standard \emph{bidirectional type system} \cite{dunfield2019,pierce2000}, which employs
two mutually defined judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
establishes that, under the typing context $\ctx$, the expression $e$ synthesizes or locally infers the type $\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$,
states that the expression $\EMV$ may appear where an expression of type $\TMV$ is expected.

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. Analysis employs the judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$,
which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini2016}, i.e. it may be
considered an arrow type. Defined in \cref{fig:calculus-matched-arrow}, this notion is purely a
technical mechanism to avoid duplication of rules related to arrow types \cite{siek2015}.

@@

We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of Hazel \cite{hazel},
a live functional programming environment in which incomplete programs, i.e., programs that contain holes, may be type-checked, manipulated, and even executed.
In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above).
Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{vouillon2014} for use in the browser.
The supplemental material contains a pre-built implementation.

The result of this effort is the first full-scale typed functional language equipped with a language server that solves the semantic gap problem: all of Hazel's semantic services are available as long as the program sketch is syntactically well-formed, which include type hints \cite{potter2020}, semantic navigation, semantic code completion \cite{potter2020,blinn2022}, contextualized documentation \cite{potter2022}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{omar2019}, even semantic services that require run-time information, like testing.
%

Hazel offers both a textual syntax, with explicit empty holes, and more interestingly,
a structure editor that is able to additionally offer partial syntactic error recovery.
In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automatically insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{moon2022,moon2023}. Efforts outside of the scope of this paper are underway to define syntactic error recovery mechanisms that can handle unmatched delimiters as well, which would achieve total syntax error recovery. In combination with our contributions, this would achieve the Hazel project's stated goal of ensuring that \emph{there are no meaningless editor states}.

@@

Earlier versions of Hazel achieved total type error recovery by using a term-based structure editor, which maintained delimiter matching by construction, albeit at the cost of some syntactic rigidity.
To specify such a structure editor formally, \citet{omar2017b} introduced the Hazelnut action calculus, which operates over a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor}.
The synthetic action judgment, written $\ASEAction{\ctx}{\ZMV}{\TMV}{\ZMV'}{\TMV'}{\AMV}$, performs some edit action $\AMV$ on some \emph{zippered expression} $\ZMV$ (which is an expression with a superimposed cursor, denoted $\hat{\EMV}$ in the original work) that synthesizes type $\TMV$, producing some new $\ZMV'$ that synthesizes types $\TMV'$.
Analogously, the analytic action judgment is written $\AAEAction{\ctx}{\ZMV}{\ZMV'}{\TMV}{\AMV}$.
Type and binding information is propagated to the location of the cursor during these operations, allowing them to insert the necessary error marks (i.e. non-empty holes, in Hazelnut's parlance).
When, for example, wrapping an expression whose type is inconsistent with $\TNum$ in an arithmetic operation, the calculus inserts an error mark around that operand:
%
\begin{mathpar}
    \inferrule{
        \inconsistent{\TMV}{\TNum}
    }{
        \ASCon{\ctx}{\ZCursor{\EMV}}{\TMV}{\ZPlusL{\ECInconType{\ECMV}}{\ZCursor{\EEHole}}}{\TNum}{\SPlusL}
    }
\end{mathpar}

### 2023-10-27 intro.tex
Clean up references.bib

Modern programming environments provide developers with a collection of semantic services%
---for example, type hints, semantic navigation, semantic code completion, and automated refactorings---%
that require static reasoning about the type and binding structure of a program as it is being edited. 
The problem is that when the program being edited is ill-typed, 
these semantic services can become degraded or unavailable \cite{HazelnutSNAPL}. 
These gaps in service are not always transient. 
For example, a change to a type definition might result in type errors at dozens of use sites in a large program, which might take hours or days to resolve, all without the full aid of these services.

These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
e.g. in the tradition of the typed lambda calculus and its derivatives \cite{TaplBook},
assigns meaning only to well-typed programs. 
If a type error appears \emph{anywhere}, the program is formally meaningless \emph{everywhere}.

This gap problem has prompted considerable practical interest in 
(1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that explain a type error, and 
(2) \textbf{type error recovery}: mechanisms that allow the system to optimistically recover from a localized type error 
and continue on to locate other errors and provide downstream semantic services, 
ideally at every location in the program and with minimal degradation in service.
Essentially all widely-used programming systems have some support for type error localization, 
e.g. in compiler error messages 
or directly in the editor via markings decorating localized errors. Developers are known to attend to reported error locations when debugging type errors \cite{DBLP:journals/jfp/JoostenBH93}.
Many systems also attempt recovery in certain situations, discussed below. 
However, type error localization and recovery mechanisms have developed idiosyncratically, 
in part as folklore amongst language and tool implementors. 
Different type checkers or language servers \cite{barros2022editing,bour2018merlin}, even for the same language, localize and recover from type errors in different ways, 
with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.

Consider, for example, the ill-typed program below, which is shown as presented to the user in
this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
\\\vspace{-5px}
\begin{center}
    \includegraphics[scale=0.5]{images/hazel-intro-screenshot-v3.png}
\end{center}

@@

An implementation with support for type error recovery, like Elm's compiler or OCaml's \li{merlin} \cite{bour2018merlin}, would 
 be tasked with continuing past this first error.
 The general difficulty here is that there is now missing semantic information, namely the type of \li{f}, that 
 the bidirectional type system as specified 
 would appear to demand in order to proceed, here in order to determine which type the argument, \li{y}, is to be analyzed against. 
 Intuitively, however, it is clear that this knowledge is \emph{unnecessary} to make an error localization decision about \li{y}: 
it is free, so a second error can be simultaneously localized despite the missing information about \li{f}.

To recover further, we might choose to also ignore the bidirectional type system's demand that the conditional guard be confirmed to be a boolean expression 
(because the type of \li{f(y)} is also not well-defined) 
and continue into its branches, observing that they have inconsistent types, \li{Bool} and \li{Int}. 
There are several ways to localize this error. 
One approach, taken e.g. by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
Heuristics have been developed to help make this choice less arbitrary, e.g. 
by observing recent editing history \cite{steady-typing}, 
or training a machine learning model \cite{SeidelBlame}. 
A less \emph{ad hoc} approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ. When the cursor is on this conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020hazel}---displays:

@@

This localization decision affects how the system recovers as it proceeds into the let body. 
If localization had assumed that the \li{then} branch were correct, as for example in \li{merlin}, then 
\li{x : Bool} and an error should be reported on only the second use of \li{x}. 
If the \li{else} branch were chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
 In either case, this error might mislead the programmer if the earlier localization guess was incorrect.
If the inconsistency were localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error at either use of \li{x}, 
so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing \cite{Siek06a}.
In any of these cases, we would like to be able to recover and 
localize the type inconsistency on the string addend because the \li{+} operator is integer addition in Hazel.
Recovering from this error, we can assume that both branch types in the second conditional will have \li{Int} type, so no 
error is marked on the conditional as a whole.

@@

\Cref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e. its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g. with destructuring patterns and System F-style polymorphism, 
that serve as case studies of the general approach that we hope will be adopted by language designers defining practical type systems.

\Cref{sec:calculus-hazel} describes our own effort in this direction, which is to scale up the marked lambda calculus as the new basis for Hazel, a typed functional programming environment that fully solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed. Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as in Agda \cite{norell:thesis}, GHC Haskell \cite{haskell-holes}, Idris \cite{DBLP:journals/jfp/Brady13}, and others), syntax error recovery \cite{medeiros20,sorkin11}, and structure editing \cite{HazelnutPOPL,teitelbaum_cornell_1981}. 
Hazel has both a textual syntax and a structure editor \cite{HazelnutPOPL,DBLP:conf/vl/Moon023}. 
%offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e. a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}. 
As a secondary contribution, 
\cref{sec:calculus-hazelnut} describes how (1)  total marking can resolve the problem of undefined behavior in the Hazelnut structure editor calculus, 
and (2) integrating marking with typed structure editing allows us to incrementally re-mark only where necessary based on the edit location and the type and binding structure.

@@

The developments in Section~\ref{sec:calculus} support the intuition (remarked upon by \citet{Localinf} and \citet{BidirTyping}) that bidirectional typing pairs well with type error localization and recovery because information flows systematically and locally through the tree. However, many functional languages including Elm, OCaml and Haskell feature constraint-based type inference, where constraints are gathered globally. For type systems where inference is decidable, this is powerful, but it is also notorious for complicating error localization, because inconsistencies can arise through a confluence of constraints originating from any number of locations \cite{DBLP:conf/popl/Wand86}.

For example, in the following Hazel expression (ignoring the error on the type hole for the moment), bidirectional error localization 
does not mark any uses of \li{x} as erroneous because a type for \li{x} cannot locally be inferred (the let-bound expression is an empty hole, which has unknown type). However, by gathering constraints from the three uses of \li{x} and attempting to unify, we see that the type hole is unfillable. Rather than arbitrarily privileging one of these uses, as is the case in languages like OCaml (and which has led to the development of complex heuristics, e.g. using machine learning \cite{SeidelBlame}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
\\
\begin{center}
    % \includegraphics[scale=0.55]{images/intro-thi.png}
    % \includegraphics[scale=0.4]{images/example_CI.png}
    \includegraphics[scale=0.4]{images/figSugg.png}
\end{center}

>>>>

Modern programming environments provide developers with a collection of semantic services%
---for example, type hints, semantic navigation, semantic code completion, and automated refactorings---%
that require static reasoning about the type and binding structure of a program as it is being edited. 
The problem is that when the program being edited is ill-typed, 
these semantic services can become degraded or unavailable \cite{omar2017a}. 
These gaps in service are not always transient. 
For example, a change to a type definition might result in type errors at dozens of use sites in a large program, which might take hours or days to resolve, all without the full aid of these services.

These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
e.g. in the tradition of the typed lambda calculus and its derivatives \cite{pierce2002},
assigns meaning only to well-typed programs. 
If a type error appears \emph{anywhere}, the program is formally meaningless \emph{everywhere}.

This gap problem has prompted considerable practical interest in 
(1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that explain a type error, and 
(2) \textbf{type error recovery}: mechanisms that allow the system to optimistically recover from a localized type error 
and continue on to locate other errors and provide downstream semantic services, 
ideally at every location in the program and with minimal degradation in service.
Essentially all widely-used programming systems have some support for type error localization, 
e.g. in compiler error messages 
or directly in the editor via markings decorating localized errors. Developers are known to attend to reported error locations when debugging type errors \cite{joosten1993}.
Many systems also attempt recovery in certain situations, discussed below. 
However, type error localization and recovery mechanisms have developed idiosyncratically, 
in part as folklore amongst language and tool implementors. 
Different type checkers or language servers \cite{barros2022,bour2018}, even for the same language, localize and recover from type errors in different ways, 
with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.

Consider, for example, the ill-typed program below, which is shown as presented to the user in
this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{czaplicki2013}. Hazel  
supports local type inference, specified in the well-established bidirectional style \cite{pierce2000,omar2017b,dunfield2019}:
\\\vspace{-5px}
\begin{center}
    \includegraphics[scale=0.5]{images/hazel-intro-screenshot-v3.png}
\end{center}

@@

An implementation with support for type error recovery, like Elm's compiler or OCaml's \li{merlin} \cite{bour2018}, would 
 be tasked with continuing past this first error.
 The general difficulty here is that there is now missing semantic information, namely the type of \li{f}, that 
 the bidirectional type system as specified 
 would appear to demand in order to proceed, here in order to determine which type the argument, \li{y}, is to be analyzed against. 
 Intuitively, however, it is clear that this knowledge is \emph{unnecessary} to make an error localization decision about \li{y}: 
it is free, so a second error can be simultaneously localized despite the missing information about \li{f}.

To recover further, we might choose to also ignore the bidirectional type system's demand that the conditional guard be confirmed to be a boolean expression 
(because the type of \li{f(y)} is also not well-defined) 
and continue into its branches, observing that they have inconsistent types, \li{Bool} and \li{Int}. 
There are several ways to localize this error. 
One approach, taken e.g. by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
Heuristics have been developed to help make this choice less arbitrary, e.g. 
by observing recent editing history \cite{chuchem2019}, 
or training a machine learning model \cite{seidel2017}. 
A less \emph{ad hoc} approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ. When the cursor is on this conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020}---displays:

@@

This localization decision affects how the system recovers as it proceeds into the let body. 
If localization had assumed that the \li{then} branch were correct, as for example in \li{merlin}, then 
\li{x : Bool} and an error should be reported on only the second use of \li{x}. 
If the \li{else} branch were chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
 In either case, this error might mislead the programmer if the earlier localization guess was incorrect.
If the inconsistency were localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error at either use of \li{x}, 
so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing \cite{siek2006}.
In any of these cases, we would like to be able to recover and 
localize the type inconsistency on the string addend because the \li{+} operator is integer addition in Hazel.
Recovering from this error, we can assume that both branch types in the second conditional will have \li{Int} type, so no 
error is marked on the conditional as a whole.

@@

\Cref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{pierce2000,dunfield2019}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{siek2006,siek2015}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{omar2017b}, operationalizing the ``red lines'' that an editor displays.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solarlezama2013}) can be marked, i.e. its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g. with destructuring patterns and System F-style polymorphism, 
that serve as case studies of the general approach that we hope will be adopted by language designers defining practical type systems.

\Cref{sec:calculus-hazel} describes our own effort in this direction, which is to scale up the marked lambda calculus as the new basis for Hazel, a typed functional programming environment that fully solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed. Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as in Agda \cite{norell2007}, GHC Haskell \cite{haskell-holes}, Idris \cite{brady2013}, and others), syntax error recovery \cite{medeiros2020,sorkin2011}, and structure editing \cite{omar2017b,teitelbaum1981}. 
Hazel has both a textual syntax and a structure editor \cite{omar2017b,moon2023}. 
%offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e. a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}. 
As a secondary contribution, 
\cref{sec:calculus-hazelnut} describes how (1)  total marking can resolve the problem of undefined behavior in the Hazelnut structure editor calculus, 
and (2) integrating marking with typed structure editing allows us to incrementally re-mark only where necessary based on the edit location and the type and binding structure.

@@

The developments in Section~\ref{sec:calculus} support the intuition (remarked upon by \citet{pierce2000} and \citet{dunfield2019}) that bidirectional typing pairs well with type error localization and recovery because information flows systematically and locally through the tree. However, many functional languages including Elm, OCaml and Haskell feature constraint-based type inference, where constraints are gathered globally. For type systems where inference is decidable, this is powerful, but it is also notorious for complicating error localization, because inconsistencies can arise through a confluence of constraints originating from any number of locations \cite{wand1986}.

For example, in the following Hazel expression (ignoring the error on the type hole for the moment), bidirectional error localization 
does not mark any uses of \li{x} as erroneous because a type for \li{x} cannot locally be inferred (the let-bound expression is an empty hole, which has unknown type). However, by gathering constraints from the three uses of \li{x} and attempting to unify, we see that the type hole is unfillable. Rather than arbitrarily privileging one of these uses, as is the case in languages like OCaml (and which has led to the development of complex heuristics, e.g. using machine learning \cite{seidel2017}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
\\
\begin{center}
    % \includegraphics[scale=0.55]{images/intro-thi.png}
    % \includegraphics[scale=0.4]{images/example_CI.png}
    \includegraphics[scale=0.4]{images/figSugg.png}
\end{center}

### 2023-10-27 polymorphism.tex
Updates from Overleaf
CCS, type classes
Clean up references.bib

We leave as future work the task of defining marked versions of even more elaborate bidirectional
type systems, e.g. \citet{dunfield2013complete}'s rather substantial formulation of implicit type
application in a bidirectional setting (which would have to handle, for example, the situation where
no implicit argument can be resolved) or \citet{DBLP:phd/hal/LennonBertrand22}'s gradual bidirectional
variant of the dependently typed calculus of constructions. Languages with operator overloading may
provide interesting localization decisions. So that overloading can be resolved during typing and
marking, the modes of one or both operand must be changed to synthesis. If the operator is not
defined given the synthesized operand type(s), an error might be marked on the operator itself.

>>>>

We leave as future work the task of defining marked versions of even more elaborate bidirectional
type systems, e.g. \citet{dunfield2013}'s rather substantial formulation of implicit type
application in a bidirectional setting (which would have to handle, for example, the situation where
no implicit argument can be resolved) or \citet{lennonbertrand2022}'s gradual bidirectional
variant of the dependently typed calculus of constructions.
Languages with operator overloading may provide interesting localization decisions.
So that overloading can be resolved during typing and marking, the modes of one or both operand must be changed to synthesis.
If the operator is not defined given the synthesized operand type(s), an error might be marked on the operator itself.
Type classes present a symmetric alternative where the operands may stay in analytic mode.
With either approach, the necessary error marks and localization decisions may be arrived at by considering the possible failure cases.
